Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-05 paper/paper.tex
related work

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model. However,
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.

@@

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
	\item Design and implementation of a probabilistic
		programming language Anglican involving tight bilateral
		integration with a general-purpose programming language.
	\item Definition of semantics of probabilistic forms in
		Anglican which facilitates full separation of
		probabilistic programs from inference algorithms.
	\item Stateless representation of exchangeable
		random procedures, or processes, seamlessly integrated
		into pure functional computation.
	\item Techniques for efficient and compact implementation of
		inference algorithms, such as representation of
		inference results as a lazy sequence of samples and 
		a novel scheme for addressing of checkpoints.
\end{itemize}

@@

@@

For inference, we chose to use  Lightweight
Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
probabilistic programming is that models are separated from
inference. To switch to a different inference algorithm we just
need to pass a different value to \texttt{doquery}. For example,
we may decide to use Black-Box Variational Bayes (BBVB), which
may not work equally well for all probabilistic programs, but is
much faster to converge.

@@

Checkpoints can be uniquely named at compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially. However, it impacts the
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item On every function call, a component is added to the
        address. Hence, the address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))]
  (sample dist))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces. However, the random values should not be
reused between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

@@

In addition to distributions, Anglican provides \textit{random processes},
which define sequences of random variables that are not independent and
identically distributed. Random processes are similar to the so called
`exchangeable random procedures' in Church \cite{GMR+08} and
Venture \cite{MSP14}. However, random sequences generated by Anglican random 
processes are not required to be exchangeable. Random processes are
defined using the \texttt{defproc} macro and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
protocol has two methods
\begin{itemize}
\item \texttt{produce}, which returns the distribution on the next
random variable in the sequence, and
\item \texttt{absorb}, which incorporates the value for the next random
variable and returns an updated random process.
\end{itemize}
 As an example, here is a definition of a beta-Bernoulli process, in which
each random variable is distributed according to a Bernoulli distribution with
an unknown parameter that is drawn from a beta distribution:
\begin{lstlisting}[style=default]
(defproc beta-bernoulli
  "beta-Bernoulli process"
  [a b]
  (produce [this] (bernoulli (/ a (+ a b))))
  (absorb [this x]
    (case x
      0 (beta-bernoulli a (inc b))
      1 (beta-bernoulli (inc a) b))))
\end{lstlisting}
Unlike typical implementations of exchangeable random processes, Anglican's random processes do not have mutable state. The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican. A sequence of random values can be generated using a recursive loop in which \texttt{absorb} returns the updated process for the next iteration. For example:
\begin{lstlisting}[style=default]
(defm sample-beta-binomial [n a b]
  (loop [process (beta-bernoulli a b)
         values []]
    (if (= (count values) n)
      values
      (let [dist (produce process)
            value (sample dist)]
      (recur (absorb process value)
             (conj values value))))))
\end{lstlisting}
Similarly, random processes can also be used to recursively observe a sequence of values:
\begin{lstlisting}[style=default]
(loop [process (beta-bernoulli 1.0 1.0)
       [value0 & values0] values] 
  (let [dist (produce process)]
    (observe dist value0)
    (recur (absorb process value0) values0)))
\end{lstlisting}

>>>>

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model. However,
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.

@@

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
	\item Design and implementation of a probabilistic
		programming language Anglican involving tight bilateral
		integration with a general-purpose programming language.
	\item Definition of semantics of probabilistic forms in
		Anglican which facilitates full separation of
		probabilistic programs from inference algorithms.
	\item Stateless representation of random processes, seamlessly integrated
		into pure functional computation.
	\item Techniques for efficient and compact implementation of
		inference algorithms, such as representation of
		inference results as a lazy sequence of samples and 
		a novel scheme for addressing of checkpoints.
\end{itemize}

@@

Efficient implementation of expressive probabilistic programming languages has
recently been an active idea of research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
SDT14,WVM14,GS15,GSX+16}. There is often a compromise between the
expressiveness of the language and the efficiency of inference.  Some languages
emphasize expressiveness~\cite{MMR+07,GMR+08,MSP14,GS15}; others restrict the
class of models which can be expressed by the language~\cite{MWG+10,SDT14,NN14}
to facilitate application of efficient inference algorithms. Anglican allows
unrestricted specification of probabilistic models in the spirit of
Church~\cite{GMR+08}, while still supporting efficient scalable
inference~\cite{WVM14,PWD+14,RNL+2016}.

Probabilistic programming languages are implemented as
interpreters~\cite{GMR+08,MSP14}, embedded languages~\cite{P09,SGG15,GSX+16},
compiled languages~\cite{MMR+07,SDT14}, as through source-to-source
transformation with augmentation~\cite{WSG11,GS15}. Each of these methods
emphasizes different priorities in language design, such as computational
efficiency, integration with an existing development environment, ease of
implementation, or efficiency of inference. Anglican implemented using a
combination of embedding and source-to-source transformational compilation to
combine advantages of both approaches.  CPS transformation, employed by
Anglican compiler, is also used in WebPPL~\cite{GS15} and facilitates clean
separation between probabilistic programs and inference algorithms and
diversity of applicable inference algorithms.

Probabilistic programming languages are often implemented as pure functional
languages~\cite{GMR+08,MSP14,SGG15}. This is because reasoning about
probabilistic programsand distribution defined by the programs is easier in
pure functional setting. Anglican is also implemented as a language without
side-effects. However, there are also languages which follow imperative
paradigm and support mutable state directly~\cite{SDT14,GSX+16}, or through
probabilistic primitives~\cite{GMR+08,MSP14}. Anglican introduces a functional
alternative to stateful random primitives in the form of random processes.

@@

For inference, we chose to use  Lightweight
Metropolis-Hastings, perhaps somewhat arbitrarily. A strength of
probabilistic programming is that models are separated from
inference. To switch to a different inference algorithm we just
need to pass a different value to \texttt{doquery}. For example,
we may decide to use Black-Box Variational Bayes (BBVB), which
may not work equally well for all probabilistic programs, but is
much faster to converge.

@@

Checkpoints can be uniquely named at compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple times due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially. However, it impacts the
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item On every function call, a component is added to the
        address. Hence, the address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))]
  (sample dist))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces. However, the random values should not be
reused between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

@@

In addition to distributions, Anglican provides \textit{random processes},
which define sequences of random variables that are not independent and
identically distributed. Random processes are similar to the so called
`exchangeable random primitives' in Church \cite{GMR+08} and
Venture \cite{MSP14}. However, random sequences generated by Anglican random 
processes are not required to be exchangeable. Random processes are
defined using the \texttt{defproc} macro and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
protocol has two methods
\begin{itemize}
\item \texttt{produce}, which returns the distribution on the next
random variable in the sequence, and
\item \texttt{absorb}, which incorporates the value for the next random
variable and returns an updated random process.
\end{itemize}
 As an example, here is a definition of a beta-Bernoulli process, in which
each random variable is distributed according to a Bernoulli distribution with
an unknown parameter that is drawn from a beta distribution:
\begin{lstlisting}[style=default]
(defproc beta-bernoulli
  "beta-Bernoulli process"
  [a b]
  (produce [this] (bernoulli (/ a (+ a b))))
  (absorb [this x]
    (case x
      0 (beta-bernoulli a (inc b))
      1 (beta-bernoulli (inc a) b))))
\end{lstlisting}
Unlike typical implementations of exchangeable random processes, Anglican's random processes do not have mutable state. The \texttt{produce} and \texttt{absorb} methods are deterministic and functional, and therefore do not have corresponding special forms in Anglican. A sequence of random values can be generated using a recursive loop in which \texttt{absorb} returns the updated process for the next iteration. For example:
\begin{lstlisting}[style=default]
(defm sample-beta-binomial [n a b]
  (loop [process (beta-bernoulli a b)
         values []]
    (if (= (count values) n)
      values
      (let [dist (produce process)
            value (sample dist)]
      (recur (absorb process value)
             (conj values value))))))
\end{lstlisting}
Similarly, random processes can also be used to recursively observe a sequence of values:
\begin{lstlisting}[style=default]
(loop [process (beta-bernoulli 1.0 1.0)
       [value0 & values0] values] 
  (let [dist (produce process)]
    (observe dist value0)
    (recur (absorb process value0) values0)))
\end{lstlisting}
